---
title: "Assignment 4: Exploratory Data Analysis"
author: "Your Name"
date: "`r Sys.Date()`"
format: html
editor: visual
---

## Overview

Due: **Friday 9/25 11:59pm**

In this assignment, you will practice Exploratory Data Analysis (EDA) by making plots from a **Kaggle dataset of your choice**.

You will submit **at least 4 different plot types**, with short written interpretations for each.

Part 1 is worth a total of 1 point, and Part 2 is worth 4 points (1 point for each plot, question, and interpretation).

## Part 1: Get Your Data

1.  Go to [Kaggle Datasets](https://www.kaggle.com/datasets), create an account, and choose a dataset that interests you. Ensure your data has:
    1.  At least 4 variables
    2.  At least one categorical and one continuous variable
    3.  I recommend data that contains at least 50-100 observations
    4.  Tip: Open the CSV on Kaggle first and skim the columns to check that it meets these requirements
2.  Download the dataset as a CSV file.
3.  Place the CSV in your class R project `data` folder.
4.  Read it into R with `read_csv()`.

```{r, message=FALSE}
library(tidyverse)
library(plotly)

avengers_data <- read_csv("data/avengers_dataset.csv")

```

### Describe your data

Data author: N/A

Link: <https://www.kaggle.com/datasets/mayankgariyamldev/avenger-data-iris-data-of-avenger-world>

Description: Data set on the five superheros in the Avengers about their power, level of their different abilities, and movie appeariencies.

### Preview your data

Use some of the strategies you've learned in this class or picked up elsewhere to summarize and quickly explore your data.

```{r}

summary(avengers_data)
```

### Summary stats

Before you make plots, you must also:

- Show basic summaries of your dataset, including:

  - `count()` for categorical variables (to see category frequencies)

  - `mean()`, `median()`, `sd()`, `min()`, `max()` for at least one numeric variable

  - Remember to use `na.rm = TRUE` if you have missing values that affect summaries!

```{r}
skim_data <- skimr::skim(avengers_data) # see what class each variable is
view(skim_data)

class(avengers_data$name)
avengers_data$name <- as.factor(avengers_data$name) # changed name from character to factor
class(avengers_data$name)
summary(avengers_data$name)

class(avengers_data$power)
avengers_data$power <- as.factor(avengers_data$power) # changed power from character to factor
class(avengers_data$power)
summary(avengers_data$power)

class(avengers_data$gender)
avengers_data$gender <- as.factor(avengers_data$gender) # changed gender from character to factor
class(avengers_data$gender)
summary(avengers_data$gender)

skim_data <- skimr::skim(avengers_data) # check the class each variable is
view(skim_data)

avengers_data |> 
  count(name, power, gender)  # to see category frequencies

avengers_data |>            # found the summaries for name
  group_by(name) |> 
  summarize(
    mean = mean(strength),
    median = median(strength),
    sd = sd(strength),
    min = min(strength),
    max = max(strength))

```

Write 2–3 sentences here: What do the category counts tell you? What does the numeric summary show about spread, center, or outliers?

The category counts tell me how many times specific categories with the same observations come up. The numeric summary tells me the average strength of the superheros. The superhero with the strongest average strength is Hulk and the superhero with the weakest average strength is Black Widow.

# Continue Here

## Part 2: Visualize Your Data

Create **at least 4 plots**, each of a different type. Your plots may share variables, but each one should answer a **different question** about your data. At least one plot must include a variable that isn't in the others.

Each plot should have:

- A title and labeled axes

- A short interpretation (2-3 sentences) underneath

  - Sample interpretation: (boxplot of `price` by `cut` from `diamonds` data): *Ideal diamonds have the lowest median price, which is surprising since Ideal is the best cut. Every cut also has a lot of very expensive outliers. One possible reason for the low Ideal prices is that Ideal diamonds might be smaller on average, which I could check by plotting carat by cut.*

Use the EDA Part 3 notes as your reference. Since this is an EDA assignment, your plots can look plain, but consider adding colors, themes, or interactivity for interpretability and visual appeal if you know how.

### Plot 1

```{r}
# insert code here
ggplot(data = avengers_data, aes(x = name, y = strength)) + 
  geom_boxplot() +
  theme_bw() +
  labs(title = "Strength of Each Avenger",
       x = "Name",
       y = "Strength")

```

Question this plot answers: Who is the strongest avenger over all the Marvel movies?

Interpretation: Over every Marvel movie, this plot shows the mean of the strength of each avenger. What is interesting is that other than the hulk, all the other avengers' strength have a big range of strength power. Although Black Widow has the lowest mean, her highest outlier of strength rivals Captain America, Ironman, and Spiderman.

### Plot 2

```{r}
# insert code here

ggplot(na.omit(avengers_data), aes(x = name, y = power)) +
  geom_point() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1, size = 5)) +
  labs(title = "With Great Power Comes Great Responsibility",
       x = "Name",
       y = "Power description") 
```

Question this plot answers: What are the different powers does each avenger have?

Interpretation: There are many different powers, abilities, and gadgets that make each person a superhero. None of the avengers have overlapping powers in this plot which was interesting to see. It shows that each superhero brings something different to the Avengers.

### Plot 3

```{r}
# insert code here
avengers_SvsA <- ggplot(data = avengers_data, aes(x = strength, y = agility)) +
  geom_jitter(alpha = 0.4, width = 0.5, color = "purple") +
  labs(title = "Strength vs Agility",
       x = "strength",
       y = "agility") +
  geom_smooth(method = "lm", se = FALSE)
  
  
ggplotly(avengers_SvsA)
```

Question this plot answers: Does more strength mean less agility and vice versa or is there no correlation?

Interpretation: Mostly when agility his the highest it has less strength and when strength is the highest it has the lowest agility. Although there are a lot of outliers, like one being 90 in strength and 99 in agility, with the line of best fit we can see that there is a correlation overall that a superhero with a higher agility has less strength and vice versa.

### Plot 4

```{r}
# insert code here
ggplot(data = avengers_data, aes(x = intelligence, y = after_stat(density))) +
  geom_freqpoly(aes(color = name), binwidth = .1, linewidth = 0.75)
```

Question this plot answers: Who is the most intelligent avenger?

Interpretation: The most intelligent avenger is Ironman. This wasn't surprising because made his own Ironman suit in a cave with a box of scraps and continued to improve the suit. After Ironman, it is close between Black Widow and Captain America and then it is Spiderman. This surprised me because Spiderman is known to be highly intelligent and was under the guidance of Ironman. This may suggest that this variable uses more than just academic intelligence although the data set doesn't specify exactly what. The least intelligent avenger is the Hulk. Which makes sense because he acts on his emotions, which is mostly anger.

------------------------------------------------------------------------

# Push to GitHub AND submit link on Canvas!!

Assignment is due Friday 9/25 at 11:59 pm.
